---
title: "Apple Game Simulation for Stockholm Bayes Club"
format:
  html:
    css: ../styles.css
    code-fold: true
    code-summary: "Show code"
---

[← Back to Teaching](../teaching.qmd){.button}

## Try one game

<button id="run-game">Play one game</button>

<div id="result-box"></div>

<style>
#run-game {
  padding: 10px 18px;
  font-size: 16px;
  border-radius: 6px;
  border: none;
  background-color: #2c7be5;
  color: white;
  cursor: pointer;
}

#run-game:hover {
  background-color: #1a5fc4;
}

#result-box {
  margin-top: 20px;
  padding: 15px;
  border-radius: 8px;
  border: 1px solid #ddd;
  max-width: 350px;
  font-size: 16px;
}

.pass {
  color: #2ca02c;
  font-weight: bold;
}

.fail {
  color: #d62728;
  font-weight: bold;
}
</style>

<script>
function simulateGame() {
  let pos = 0;
  let apples = 0;
  let skip = false;

  while (pos < 21) {

    if (skip) {
      skip = false;
      continue;
    }

    const roll = Math.floor(Math.random() * 6) + 1;

    if (roll === 1 || roll === 2 || roll === 6) {
      pos += roll;

      if (pos === 9) pos = 14;
      if (pos === 18) skip = true;

      if (pos >= 21) break;

    } else {
      apples += roll;
    }
  }

  return apples;
}

function check(value, threshold) {
  if (value >= threshold) {
    return `<span class="pass">✔ ≥ ${threshold}</span>`;
  } else {
    return `<span class="fail">✖ ≥ ${threshold}</span>`;
  }
}

document.getElementById("run-game").addEventListener("click", () => {

  const apples = simulateGame();

  document.getElementById("result-box").innerHTML =
    `<strong>Apples collected:</strong> ${apples}<br><br>
     ${check(apples,10)}<br>
     ${check(apples,20)}<br>
     ${check(apples,30)}`;
});
</script>

---

## Distribution from 10,000 simulated games

```{r}
library(ggplot2)

simulate_game <- function() {
  pos <- 0
  apples <- 0
  skip <- FALSE

  while (pos < 21) {

    if (skip) {
      skip <- FALSE
      next
    }

    roll <- sample(1:6, 1)

    if (roll %in% c(1,2,6)) {
      pos <- pos + roll

      if (pos == 9) pos <- 14
      if (pos >= 21) break
      if (pos == 18) skip <- TRUE

    } else {
      apples <- apples + roll
    }
  }

  apples
}

set.seed(123)

n_sim <- 10000
results <- replicate(n_sim, simulate_game())

df <- data.frame(apples = results)

ggplot(df, aes(x = apples)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "#E1635B",
    color = "white"
  ) +
  scale_x_continuous(breaks = seq(0, max(df$apples), by = 5)) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 6),
    labels = scales::comma
  ) +
  labs(
    title = "Distribution of Apples Collected",
    subtitle = "10,000 simulated games",
    x = "Apples collected",
    y = "Number of games"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```
## Estimated probability of winning the game for 10, 20, 30 apples.

```{r}
library(scales)

p10 <- percent(mean(results >= 10))
p20 <- percent(mean(results >= 20))
p30 <- percent(mean(results >= 30))

cat(
  " ≥10 apples:", p10,"\n",
  "≥20 apples:", p20,"\n",
  "≥30 apples:", p30
)
```

[← Back to Teaching](../teaching.qmd){.button}
